A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-18 tex/codatadp.tex
Assumptions for section 3

  * SMT background

\rem{Discuss signature $\Sig$, theory $\thD$}
\rem{DPLL(T) : we can restrict ourselves to conjunctions.}
This section presents a calculus for determining the satisfiability of conjunctions of $\thD$-constraints.

Our calculus consists of derivation rules that operate on a set $\Ec$ of equalities and disequalities between $\Sig$-terms.
We write $\tEc$ to denote the set of $\Sig$-terms occurring in $\Ec$.
%We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $\const{s}^{\,k}_j( \const{C}_j(t_1,\ldots,t_n) )$
are simplified to $t_i$, and moreover assume that all %additional
constraints added to $\Ec$ are normalized in this manner.

@@

\begin{lemma} 
\label{lem:model-completion}
If $\Val$ is constructed for a saturated $\Ec$,
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then:
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$, for all $\ec{x}$ of type $\tau$ of $\Ec$,
\item
If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
then $\Val^\ast \ec{x} = \mu v_{\ec{x}}. \const{C}( w_1, \ldots, w_n )$
and $\Val^\ast \ec{x} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%and $\muvar( w_i ) = v_{\ec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)=v_{\ec{x}}$,
\item 
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable $v_{\ec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else $v_{\ec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_x(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to $v_{\ec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction, 
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to $v_{\ec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Cyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form 
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$.

@@

\paragraph{Optimizations}
We first discuss several optimizations that are not reflected in the presentation of the calculus, which closely follow the approach outlined by Barrett et al in~\cite{}.
We only briefly mention these optimizations, since each applies to datatypes as well as co-datatype, and most have been presented thoroughly in~\cite{}.

Discriminators are treated as predicate symbols, so a discriminator $\const{d}_i( t )$ is such that $\const{d}_i$ 
is a predicate symbol, instead of reducing this constraint to $t \teq \const{C}_i( s^1_i( t ), \ldots, s^n_i( t ) )$.
We have found that this leads to better performance since this reduction introduces terms more eagerly to $\tEc$.
Handling discriminators requires extending the decision procedure with several additional rules (see~\cite{} for more details), which apply uniformly to both datatypes and co-datatypes.
Also, selectors are collapsed eagerly:
if $s^j_i( t )$ is a term in $\tEc$ and $t$ contains the constructor 
$\const{C}_i( \vec u )$, then we immediately infer $s^j_i( t ) \teq u_j$, whereas the calculus we presented would apply \rn{Split} and \rn{Decomp} before inferring this equality.  
For the sake of reducing the number of unique constraints considered by the calculus, we compute a normal form for constraints as a pre-processing step.
In particular, we rewrite $t \teq s$ to $s \teq t$ if $s$ is less than $t$ based on a term ordering,
replace $\const{C}_i( \vec t ) \teq \const{C}_j( \vec u )$ with $\bot$ when $i \neq j$,
replace all selector terms of the form $\const{s}^{\,i}_j( \const{C}_j(t_1,\ldots,t_n) )$ by $t_i$,
and replace all occurrences of discriminators $\const{d}_i( \const{C}_k( \vec t ) )$ by $\top$ if $i=k$ and $\bot$ if $i \neq k$.

>>>>

This section presents a decision procedure for \rem{introduce ``$\thD$" in previous section?}
determining the $\thD$-satisfiability of sets of constraints built from a fixed signature $\Sig$.
We will make the following assumptions about $\Sig$.
First, all codatatypes in $\Codata$ are corecursive.
This is not restrictive, since checking the $\thD$-satisfiability for constraints involving a non-corecursive co-datatype $\tau$ can be trivially reduced to
checking the satisfiability of the same constraint involving a (non-recursive) datatype of the same form as $\tau$. \rem{revise?}
Second, all types $\tau \in \Nondata$ have infinite cardinality.
This also is not restrictive, since each interpreted type $\tau \in \Nondata$ having finite cardinality $n$
can be replaced by a enumeration datatype with $n$ constructors
\footnote{Doing so simplifies the presentation of the calculus, but is not done in practice. \rem{More?}}.
Additionally, since $\Ec$ is ground, we know that $\Ec$ cannot entail an upper bound on the cardinality of any uninterpreted type $\tau \in \Nondata$, 
and thus without loss of completeness we may treat all other types as infinite.

Our calculus consists of derivation rules that operate on a set $\Ec$ of equalities and disequalities between $\Sig$-terms,
where $\Ec$ is interpreted as a conjunction.
We write $\tEc$ to denote the set of $\Sig$-terms occurring in $\Ec$.
%We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $\const{s}^{\,k}_j( \const{C}_j(t_1,\ldots,t_n) )$
are simplified to $t_i$, and moreover assume that all %additional
constraints added to $\Ec$ are normalized in this manner.

@@

\begin{lemma} 
\label{lem:model-completion}
If $\Val$ is constructed for a saturated $\Ec$,
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then:
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$, for all $\ec{x}$ of type $\tau$ of $\Ec$,
\item
If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
then $\Val^\ast \ec{x} = \mu v_{\ec{x}}. \const{C}( w_1, \ldots, w_n )$
and $\Val^\ast \ec{x} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%and $\muvar( w_i ) = v_{\ec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)=v_{\ec{x}}$,
\item 
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable $v_{\ec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else $v_{\ec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_x(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to $v_{\ec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction, 
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to $v_{\ec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Cyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form 
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$.

@@

\paragraph{Optimizations}
We first discuss several optimizations that are not reflected in the presentation of the calculus, which closely follow the approach outlined by Barrett et al in~\cite{}.
We only briefly mention these optimizations, since each applies to datatypes as well as codatatypes, and most have been presented thoroughly in~\cite{}.

Discriminators are treated as predicate symbols, so a discriminator $\const{d}_i( t )$ is such that $\const{d}_i$ 
is a predicate symbol, instead of reducing this constraint to $t \teq \const{C}_i( s^1_i( t ), \ldots, s^n_i( t ) )$.
We have found that this leads to better performance since this reduction introduces terms more eagerly to $\tEc$.
Handling discriminators requires extending the decision procedure with several additional rules (see~\cite{} for more details), which apply uniformly to both datatypes and codatatypes.
Also, selectors are collapsed eagerly:
if $s^j_i( t )$ is a term in $\tEc$ and $t$ contains the constructor 
$\const{C}_i( \vec u )$, then we immediately infer $s^j_i( t ) \teq u_j$, whereas the calculus we presented would apply \rn{Split} and \rn{Decomp} before inferring this equality.  
For the sake of reducing the number of unique constraints considered by the calculus, we compute a normal form for constraints as a pre-processing step.
In particular, we rewrite $t \teq s$ to $s \teq t$ if $s$ is less than $t$ based on a term ordering,
replace $\const{C}_i( \vec t ) \teq \const{C}_j( \vec u )$ with $\bot$ when $i \neq j$,
replace all selector terms of the form $\const{s}^{\,i}_j( \const{C}_j(t_1,\ldots,t_n) )$ by $t_i$,
and replace all occurrences of discriminators $\const{d}_i( \const{C}_k( \vec t ) )$ by $\top$ if $i=k$ and $\bot$ if $i \neq k$.
